---
title: "Survival - Practical 4: Cox PH Models Extensions"
output: 
  learnr::tutorial:
     allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library("learnr")
library("survival")
library("lattice")
library("splines")
data("lung", package = "survival")
lung$sex <- factor(lung$sex, labels = c("male", "female"))
lung <- with(lung, lung[complete.cases(time, status, sex, age, ph.karno), ])
knitr::opts_chunk$set(echo = FALSE)
```


## Quiz

The following questions test your knowledge in the extensions of Cox proportional hazards 
models, presented in Chapter 5.

### Question 1

```{r Q1}
quiz(
  question("Why do we need the Breslow estimator to calculate survival probabilities from a Cox model (more than one correct possible)?",
    answer("Because the Cox model does not estimate the baseline hazard.", correct = TRUE),
    answer("Because the Cox model does not make an assumption for the distribution of the survival times.", correct = TRUE),
    answer("Because the Cox model makes the proportional hazards assumption."),
    answer("Because the Cox model does not make an assumption for the form of the baseline hazard.", correct = TRUE),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

### Question 2

```{r Q2}
quiz(
  question("Which of the following statements are reasons to use a stratified Cox model (more than one correct possible)?",
    answer("To account for categorical covariates that do not satisfy PH.", correct = TRUE),
    answer("To account for continuous covariates that do not satisfy PH."),
    answer("To account for categorical covariates that are time-varying."),
    answer("To account for continuous covariates that are time-varying"),
    answer("To account for heterogeneity caused by categorical covariates.", correct = TRUE),
    answer("To account for heterogeneity caused by continuous covariates."),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

### Question 3

```{r Q3}
quiz(
  question("Which of the following examples are endogenous time-varying covariates (more than one correct possible)?",
    answer("Prostate-specific antigen levels for prostate cancer patients.", correct = TRUE),
    answer("Type of nurse that treats the patients."),
    answer("Type of hospital (university or not) visited by the patient."),
    answer("Aortic gradient values of patient who received an aortic valve transplantation.", correct = TRUE),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

### Question 4
A randomized study has been performed to assess the efficacy and safety of a new drug for 
AIDS. Patients were randomized either to the new drug or to the standard treatment, and 
the primary outcome was time to death. We have collected data from 55 different centers 
and we would like to test whether the new treatment significantly prolongs survival over 
all the centers (i.e., we want the pooled effect). Relatively few deaths have been 
recorded during follow-up, and in some centers we did not have any event.

```{r Q4}
quiz(
  question("Which of the following types of analysis would you follow?",
    answer("A Cox model that includes as covariate the treatment indicator."),
    answer("A Cox model that includes as covariate the treatment indicator, but with corrected standard errors taking into account that we have data for different centers.", correct = TRUE),
    answer("A Cox model that includes as covariate the treatment indicator, and a frailty term for center."),
    answer("A Cox model that includes as covariates the treatment indicator and the center indicator."),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

### Question 5

In a study on HIV infected patients, we are interested in the time until they develop AIDS.
However, some patients die before an AIDS diagnosis for reasons related to their 
disease. We would like to estimate the survival function for the time to AIDS, while 
accounting for these deaths. 

```{r Q5}
quiz(
  question("Which of the following options would you choose?",
    answer("A non-parametric estimator of the cumulative incidence functions for the time to AIDS accounting for competing risks.", correct = TRUE),
    answer("The Kaplan-Meier estimator of the survival function for the time to AIDS, treating patient who died as censored."),
    answer("Fit a Cox model for the time to AIDS, treating patients who died as censored. Use the Breslow estimator to derive an estimate of the survival function for the time to AIDS."),
    answer("Fit an AFT model for the time to AIDS, treating patients who died as censored. Use the Breslow estimator to derive an estimate of the survival function for the time to AIDS."),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

## Exercises

The purpose of this exercise is to illustrate how to perform a representative Cox PH regression 
analysis, including the extensions seen in the last sections of Chapter 4 and in Chapter 5.

The following questions are based on the Lung data set. This data set is available as 
the object `lung` from package **survival**, which is already loaded in this session. 
From this data set we will use the following variables:

- `time`: the observed time-to-death in days.

- `status`: the event indicator; '1' denotes censored and '2' denotes death.

- `age`: the age in years.

- `ph.karno`: Karnofsky performance score rated by the physician.

- `sex`: the sex indicator with values 'male' and 'female'.

- `ph.ecog`: ECOG performance score (0=good 5=dead).

For the exercises below it will be useful to check the corresponding sections of the
[Survival Analysis in R Companion](http://www.drizopoulos.com/courses/EMC/EP03_%20Survival%20Analysis%20in%20R%20Companion.html) that are given in the hints.

### Question 1

Our initial hypothesis is that the time-to-death is affected by `sex`, `age` and 
`ph.karno`. In addition, the physicians believe that the effect of `ph.karno` and `age` 
may be nonlinear in the log-hazard scale. Moreover, the (possibly nonlinear -- model using
natural cubic splines with 3 degrees of freedom) effects of  `age` and `ph.karno` on the 
log-hazard scale are not the same for males and females. Transform this initial hypothesis
into a suitable Cox PH model. Name the fitted model `fit1`.

The aim here is to do a realistic analysis of a survival dataset with a Cox PH model. Hence,

a. Start by checking if the interaction effects can be dropped using a likelihood ratio test.

b. Similarly check if the nonlinear effects can be dropped using a likelihood ratio test.

c. For the final model check the PH assumption for the terms in the model.

```{r P1, exercise=TRUE, exercise.lines = 20}

```

```{r P1-hint-1}
# Check the examples in slides 191 & 262-263, and 
# Section 4.1, Survival Analysis in R Companion

# Fit the Cox PH model with nonlinear and interaction effects
```
```{r P1-hint-2}
# The code is (this code only fits the model and does not produce any output):
fit1 <- coxph(Surv(time, status) ~ sex * (ns(age, 3) + ns(ph.karno, 3)), 
              data = lung)
```

```{r P1-hint-3}
# Fit the Cox PH model without the interaction effects, and do the LRT
# using anova()
```
```{r P1-hint-4}
# The code is
fit2 <- coxph(Surv(time, status) ~ sex + ns(age, 3) + ns(ph.karno, 3), 
              data = lung)

anova(fit2, fit1)
```

```{r P1-hint-5}
# Fit the Cox PH model without the nonlinear effects, and do the LRT.
# Compare this model versus the full model that you initially fitted
```
```{r P1-hint-6}
# The code is
fit3 <- coxph(Surv(time, status) ~ sex + age + ph.karno, 
              data = lung)

anova(fit3, fit2)
```

```{r P1-hint-7}
# Based on the results of the previous two tests decide how to simplify the model.

# Fit the final model fit4
```
```{r P1-hint-8}
# The final model is (this code only fits the model and does not produce any output):
fit4 <- coxph(Surv(time, status) ~ sex + age + ph.karno, 
              data = lung)
```

```{r P1-hint-9}
# Check the PH assumption using function cox.zph() - check the code in slide 240, and 
# Section 4.4, Survival Analysis in R Companion

# Produce the plot separately per variables and add the coefficient from the final Cox 
# model (i.e., the PH assumption). We want to see if the red line is a good 
# approximation of the black one. Check Section 4.4, Survival Analysis in R Companion
```
```{r P1-hint-10}
# The code is
check_PH <- cox.zph(fit4)

plot(check_PH, var = 1)
abline(h = coef(fit4)[1], col = "red", lwd = 2)

plot(check_PH, var = 2)
abline(h = coef(fit4)[2], col = "red", lwd = 2)

plot(check_PH, var = 3)
abline(h = coef(fit4)[3], col = "red", lwd = 2)
```

```{r P1-hint-11}
# We observe that the PH assumptions does not seem to be satisfied for 'age' and
# 'ph.karno'.

# One solution discussed in slide 243 is to partion the time axis. From oberve that 
# from the two variables, for 'age' the PH seems to be more strongly violated. In addition,
# from the plot we observe that there are potentially three periods in which 
# the coefficient of age seems more constant, namely in the first 100 days, then 
# from 100 to 270 days, and finally from 270 days until the end.
# Hence, we decide to create a partion of the complete dataset in these three pieces.
# Then we will fit a Cox model that will allow for different coefficients of 'age' and
# 'ph.karno' in the three partitions.

# We start by creating the partioned dataset. This can be easily done using the function
# survSplit() from the survival package. Check the help file and the example for this 
# function. We will need to give the formula which is the same as the formula of model
# 'fit4', provide the original 'lung' data in the 'data' argument, then define in the 'cut' 
# argument the partition, i.e., give the vector c(100, 270), and finaly also define the 
# name of the variable that will denote the different pieces of the data; name this 
# variable 'time_group'

# Create the partioned dataset using survSplit() as described above, and name it 'lung2'.
# Check Section 5.2.1, Survival Analysis in R Companion
```
```{r P1-hint-12}
# The code is:
lung2 <- survSplit(Surv(time, status) ~ sex + age + ph.karno, data = lung,
                   cut = c(100, 270), episode = "time_group")
```

```{r P1-hint-13}
# Fit a Cox model in the lung2 dataset that includes as stratifying factor the 
# 'time_group' variable and its interaction with 'age' and 'ph.karno'.
# Name the fitted model 'fit5'
```
```{r P1-hint-14}
# The code is:
fit5 <- coxph(Surv(tstart, time, status) ~ sex + (age + ph.karno):strata(time_group),
              data = lung2)
```

```{r P1-hint-15}
# Check the PH assumption, as before using cox.zph(), for the first model.
# Do also the plots.
```
```{r P1-hint-16}
# The code is:

# First we define a helper function to do the plot and add the horizontal line
plt_zph <- function (object, var) {
    plot(cox.zph(object), var = var)
    abline(h = coef(object)[var], col = "red", lwd = 2)
}

# check PH for sex
plt_zph(fit5, var = 1)

# check PH for age
opar <- par(no.readonly = TRUE, mfrow = c(2, 2), oma = c(3, 3, 2, 3), 
            mar = c(2, 0, 0, 0), mgp = c(3, 0.4, 0), tcl = -0.25)
plt_zph(fit5, var = 2)
plt_zph(fit5, var = 3)
plt_zph(fit5, var = 4)
par(opar)

# check PH for ph.karno
opar <- par(no.readonly = TRUE, mfrow = c(2, 2), oma = c(3, 3, 2, 3), 
            mar = c(2, 0, 0, 0), mgp = c(3, 0.4, 0), tcl = -0.25)
plt_zph(fit5, var = 5)
plt_zph(fit5, var = 6)
plt_zph(fit5, var = 7)
par(opar)
```

### Question 2

We are interested in estimating survival probabilities for males and females with the 
median age, and with the average Karnofsky score. 

a. Which are the median survival times and their 95% confidence limits for males and 
females with median age and average Karnofsky score?

b. Plot the corresponding survival curves.

c. What are the corresponding survival probabilities for 200, 400, 600 and 800 days?

```{r P2-setup}
lung2 <- survSplit(Surv(time, status) ~ sex + age + ph.karno, data = lung,
                   cut = c(100, 270), episode = "time_group")
fit5 <- coxph(Surv(tstart, time, status) ~ sex + (age + ph.karno):strata(time_group),
              data = lung2)

summary(fit5)
```

```{r P2, exercise=TRUE, exercise.lines = 20}

```

```{r P2-hint-1}
# Remember that we work with the statified Cox model 'fit5' and the paritioned dataset
# at 100 and 270 days.

# First we need to define the dataset that contains the information of the type of 
# patients for which we want to calculate the survival probabilities.

# Create this dataset using as source (i.e., to calculate the median age and mean
# Karnofsky score) the original 'lung' dataset.

# Use expand.grid() - check the code in slides 292-293, and 
# Section 5.1, Survival Analysis in R Companion
```
```{r P2-hint-2}
# The code is (this code only creates the dataset and does not produce any output):
ND <- with(lung, expand.grid(sex = levels(sex), age = median(age), 
                             ph.karno = mean(ph.karno), time_group = 1:3))
```

```{r P2-hint-3}
# Calculate the survival probabilities using this dataset 'ND' and the statified Cox 
# model 'fit5' using the survfit() function.
```
```{r P2-hint-4}
# The code is:
probs <- survfit(fit5, newdata = ND)

probs

plot(probs)
```

```{r P2-hint-5}
# We want the survival probabilities at 200 days. For this we will need to use the second 
# part of the patition from 100 to 270 days. 

# Then using the summary() method and the 'times' argument.
```
```{r P2-hint-6}
# The code is:
probs_200 <- survfit(fit5, newdata = ND[ND$time_group == 2, ])

summary(probs_200, times = 200)
```


```{r P2-hint-7}
# Likewise to calculate the survival probabilities at 400, 600 and 800 days we need the
# third part of the partition from 270 days to the end of the study.

# Then using the summary() method and the 'times' argument.
```
```{r P2-hint-8}
# The code is:
probs_400 <- survfit(fit5, newdata = ND[ND$time_group == 3, ])

summary(probs_400, times = c(400, 600, 800))
```

### Question 3

For the rest of the questions we consider the additive Cox PH model with `sex`, `age` and 
`ph.karno` fitted in the original `lung` database (i.e., not the two databases before and 
after 170 days). It is believed that the baseline hazard of death has a completely 
different shape for patients with ECOG score greater than 0 compared to patients with ECOG
equal to 0, i.e., the hazard functions of the two groups is not analogous. First, from the
`ph.ecog` variable that takes values from 0 to 3, construct the variable `ph.ecog2` that 
groups together the values 1-3. Then, fit an appropriate Cox model that takes the feature 
described above into account, and then interpret the results. Name the fitted model 
`fit6`.

```{r P3, exercise=TRUE, exercise.lines = 15}

```

```{r P3-hint-1}
# First, we need to construct the 'ph.ecog2'. Start by setting the variable as a copy of
# the original 'ph.ecog'. Then set the values of 'ph.ecog2' which are greater than 0 
# equal to 1.
```
```{r P3-hint-2}
# The code is (this code only constructs the variable and does not produce any output):
lung$ph.ecog2 <- lung$ph.ecog
lung$ph.ecog2[lung$ph.ecog2 > 0] <- 1
```

```{r P3-hint-3}
# First, we need to construct the 'ph.ecog2'. Start by setting the variable as a copy of
# the original 'ph.ecog'. Then set the values of 'ph.ecog2' which are greater than 0 
# equal to 1.
```
```{r P3-hint-4}
# The code is:
fit6 <- coxph(Surv(time, status) ~ sex + age + ph.karno + strata(ph.ecog2), data = lung)

summary(fit6)
```

### Question 4

The team of physicians of the North Central Cancer Treatment Group (who are responsible
for the Lung study) believe that the effects of `sex`, `age` and `ph.karno` in the risk 
of death are different for the two ECOG groups. Extend the model `fit6` of Q3 accordingly,
and test whether this hypothesis is supported by the data for each of the two predictors.

```{r P4-setup}
lung$ph.ecog2 <- lung$ph.ecog
lung$ph.ecog2[lung$ph.ecog2 > 0] <- 1
fit6 <- coxph(Surv(time, status) ~ sex + age + ph.karno + strata(ph.ecog2), data = lung)
```
```{r P4, exercise=TRUE, exercise.lines = 15}

```

```{r P4-hint-1}
# Check slides 299 & 302, and Section 5.2, Survival Analysis in R Companion

# You will need to include the interaction between the stratifying factor and the three
# covariates, and do the LRT to compare with the previous model.
```
```{r P4-hint-2}
# The code is:
fit7 <- coxph(Surv(time, status) ~ (sex + age + ph.karno) * strata(ph.ecog2), data = lung)

anova(fit6, fit7)
```
