---
title: "Survival - Practical 3: Cox PH Models"
output: 
  learnr::tutorial:
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library("learnr")
library("survival")
library("lattice")
library("splines")
load("Data.RData")
knitr::opts_chunk$set(echo = FALSE)
options(tutorial.exercise.timelimit = 1200)
```


## Quiz

The following questions test your knowledge in the framework of Cox proportional hazards 
models, presented in Chapter 4.

### Question 1

```{r Q1}
quiz(
  question("In Cox PH models the regression coefficients in their original scale (i.e., no transformation) quantify changes for which quantity?",
    answer("The log failure time."),
    answer("The failure time."),
    answer("The expected log failure time."),
    answer("The expected failure time."),
    answer("The log hazard function.", correct = TRUE),
    answer("The expected log hazard function."),
    answer("The hazard function."),
    answer("None of the above."),
    allow_retry = TRUE, random_answer_order = FALSE
  )
)
```

### Question 2
In a study for hypertensive patients, a new medical treatment is compared with the 
traditional medication. The hazard ratio for a major cardiac incident between the two
treatments is 0.78 (95% CI: 0.71; 0.86).

```{r Q2}
quiz(
  question("Which of the following statements is correct?",
    answer("The new treatment lowers the chance of a major cardiac event by 22%."),
    answer("The new treatment lowers the risk of a major cardiac event by 22%."),
    answer("The traditional treatment increases the instaneous risk of a major cardiac event by 28%.", correct = TRUE),
    answer("The traditional treatment increases the instaneous risk of a major cardiac event by 22%."),
    answer("The traditional treatment increases the risk of a major cardiac event by 28%."),
    answer("The traditional treatment increases the risk of a major cardiac event by 22%."),
    answer("None of the above statements is correct."),
    allow_retry = TRUE, random_answer_order = FALSE
  )
)
```

### Question 3
Continuing on the result from the study described in Question 2, we observe that the 95%
confidence interval for the hazard ratio is slightly non-symetric.

```{r Q3}
quiz(
  question("Which of the following statements is correct?",
    answer("The 95% CI is non-symmetric due to rounding error."),
    answer("The 95% CI has been wrongly calculated, it should be exactly symmetric."),
    answer("The 95% CI for the log hazard ratio is symmetric, not for the hazad ratio.", correct = TRUE),
    answer("The 95% CI is non-symmetric due to the semi-parametric nature of the Cox model (i.e., as in other non-parametric methods)."),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

### Question 4

```{r Q4}
quiz(
  question("Why Cox PH models are called semi-parametric models?",
    answer("Because they make no assumptions for the distribution of the covariates in the model."),
    answer("Because they make no assumptions for the distribution of the survival times.", correct = TRUE),
    answer("Because the effects of covariates on the hazard function can be nonlinear."),
    answer("Because they do not have error terms as AFT models do."),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

### Question 5

```{r Q5}
quiz(
  question("What is a consequence of the fact that Cox models are semi-parametric (more than one correct possible)?",
    answer("We cannot obtain absolute instantaneous risks only relative instantaneous risks.", correct = TRUE),
    answer("We cannot obtain relative instantaneous risks only absolute instantaneous risks."),
    answer("We cannot obtain survival probabilities.", correct = TRUE),
    answer("We cannot obtain expected failure times as in AFT models.", correct = TRUE),
    answer("We cannot obtain relative risks (i.e., ratio of probabilities not hazards).", correct = TRUE),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

### Question 6

```{r Q6}
quiz(
  question("Which of the following statements stems from the proportional hazards assumption?",
    answer("The instantaneous risk of an event remains constant over time."),
    answer("The effect of covariates on the hazard remains constant over time.", correct = TRUE),
    answer("The baseline instantaneous risk remains constant over time."),
    answer("Covariates remain constant over time (i.e., we have only baseline covariates)."),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

## Exercises

The purpose of this practical is to illustrate how the Cox proportional hazards model can 
be fitted in R.

The following questions are based on the AIDS dataset. This dataset is available as the 
object `aids.id` and is already loaded in this session. From this dataset we will use the 
following variables:

- `Time`: the observed time-to-death in months.

- `death`: the event indicator; '1' denotes death and '0' censored observation.

- `CD4`: baseline CD4 cell count measurement.

- `drug`: the treatment indicator with values 'ddC' and 'ddI'.

- `AZT`: indicator denoting whether the patient was enrolled because of AZT 'intolerance' 
or AZT 'failure'.

For the exercises below it will be useful to check the corresponding sections of the
[Survival Analysis in R Companion](http://www.drizopoulos.com/courses/EMC/EP03_%20Survival%20Analysis%20in%20R%20Companion.html) that are given in the hints.

### Question 1

Fit a Cox model that relaxes the linearity assumption for the effect of `CD4` using 
natural cubic splines (for this you will need to use function `ns()`, i.e., within the 
formula argument of `coxph()` use `ns(CD4, 3)`). In addition, include the main effects of 
`drug` and `AZT`, and the interaction effects of `CD4` with both `drug` and `AZT`. Name 
the fitted model `fit1`.

```{r P1, exercise=TRUE, exercise.lines = 15}

```

```{r P1-hint-1}
# Check the examples in slides 191 & 262-263, and 
# Section 4.1, Survival Analysis in R Companion 
```
```{r P1-hint-2}
# Fit the Cox model
fit1 <- coxph(Surv(Time, death) ~ ns(CD4, 3) * (drug + AZT), data = aids.id)
summary(fit1)
```

### Question 2

Use a likelihood ratio test to test whether model `fit1` can be reduced by dropping all 
interaction terms. Depending on the result choose the model that you will use for the 
remaining questions unless otherwise stated. Name the final fitted model `fit2`.

```{r P2-setup}
fit1 <- coxph(Surv(Time, death) ~ ns(CD4, 3) * (drug + AZT), data = aids.id)
```

```{r P2, exercise=TRUE, exercise.lines = 10}

```

```{r P2-hint-1}
# Check the example in slides 213, and Section 4.3, Survival Analysis in R Companion
```
```{r P2-hint-2}
# We first fit the model under the null hypothesis (i.e., removing the interaction terms)
fit2 <- coxph(Surv(Time, death) ~ ns(CD4, 3) + drug + AZT, data = aids.id)

# Then using anova() we compare with fit1 using a likelihood ratio test
anova(fit2, fit1)
```

### Question 3

Use the `summary()` method to obtain a detailed summary of the fitted model `fit2`. What 
is the interpretation of the estimated coefficient for `drug`? In addition, in the output 
you have values for `exp(coef)` and `exp(-coef)`. What do these values represent? 

```{r P3-setup}
fit2 <- coxph(Surv(Time, death) ~ ns(CD4, 3) + drug + AZT, data = aids.id)
```

```{r P3, exercise=TRUE, exercise.lines = 10}

```

```{r P3-hint-1}
# Just use summary() for model fit2.
```
```{r P3-hint-2}
summary(fit2)
```

### Question 4

Using the model of Question 1 ,`fit1`, create an effect plot depicting how the average log
hazard ratio changes with increasing values of `CD4`, for 'ddI' and 'ddC' patients who had 
enrolled because of either AZT 'intolerance' or AZT 'failure'. What do you observe?

```{r P4-setup}
fit1 <- coxph(Surv(Time, death) ~ ns(CD4, 3) * (drug + AZT), data = aids.id)
```

```{r P4, exercise=TRUE, exercise.lines = 15}

```

```{r P4-hint-1}
# Check the example in slides 198-201, and Section 4.2, Survival Analysis in R Companion
```

```{r P4-hint-2}
# We create the database based on which predictions will be made. We need three 
# variables/columns: 'drug' (both 'ddC' and 'ddI'), 'AZT' (both 'intolerance' and 'failure'), 
# and 'CD4' increasing from 0 to 10.

# Use function expand.grid()
```

```{r P4-hint-3}
# The code is (this code just creates the dataset and does not create any output)
ND <- expand.grid(CD4 = seq(0, 10, length.out = 20), drug = c("ddC", "ddI"), 
                  AZT = c("intolerance", "failure"))

```

```{r P4-hint-4}
# Calculate the predictions using the same code as in slide 200, using model 'fit1'
```

```{r P4-hint-5}
# The code is
prs <- predict(fit1, newdata = ND, type = "lp", se.fit = TRUE)
ND$pred <- prs[[1]]
ND$se <- prs[[2]]
ND$lo <- ND$pred - 1.96 * ND$se
ND$up <- ND$pred + 1.96 * ND$se

# The first lines of ND are
head(ND)
```

```{r P4-hint-6}
# Produce the plot using the same code as in slide 201.
# Note, you will need to adapt a bit this code, namely,
# you will need to use 'CD4' instead of 'age', and 
# 'AZT*drug' instead of  only 'drug'
```

```{r P4-hint-7}
# The code is
xyplot(pred + lo + up ~ CD4 | AZT * drug, data = ND, 
       col = c("red", "black", "black"), lty = c(1, 2, 2), lwd = 2, type = "l",
       abline = list(h = 0, lty = 2), xlab = "CD4", ylab = "log Hazard Ratio")
```

### Question 5

Using the Kaplan-Meier estimator to compare whether the proportional hazards assumption is 
justified for `AZT`.

```{r P5, exercise=TRUE, exercise.lines = 15}

```

```{r P5-hint-1}
# Check the example in slide 229, and Section 4.4, Survival Analysis in R Companion
```

```{r P5-hint-2}
# First calculate the Kaplan-Meier estimator for the two levels of AZT;
# this is done using survfit()
```

```{r P5-hint-3}
# The code is (this code just computes the KM estimator and does not create any output)
fit <- survfit(Surv(Time, death) ~ AZT, data = aids.id)
```

```{r P5-hint-4}
# Then produce the figure using the code from slide 229
```

```{r P5-hint-5}
# The code is
par(mfrow = c(1, 2))
plot(fit, xlab = "Months", ylab = "Survival", col = 1:2)
plot(fit, fun = function (s) -log(-log(s)), xlab = "Months", 
    ylab = "-log(- log(Survival))", col = 1:2)
```
